---
title: "Treatments and timepoints"
author: "Filippo Biscarini"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
library("DT")
library("broom")
library("knitr")
library("ggpubr")
library("contrast")
library("tidyverse")
library("data.table")

knitr::opts_chunk$set(echo = TRUE)
```

## Treatments and timepoints

$$
y = \mu + \beta_1 \text{timepoint} + \beta_2 \text{treatment} + e
$$

## Read the data

Data from [The effects of lateral line ablation and regeneration in schooling giant danios](https://journals.biologists.com/jeb/article/221/8/jeb175166/300/The-effects-of-lateral-line-ablation-and) (data repo [here](https://zenodo.org/record/4999506))

It's data on fish (*Devario aequipinnatus*) lateral line system, with the effect of chemical treatments (gentamycin or no treatment -- sham) at different timepoints.

```{r read-the-bird-data}
basefolder = "/home/filippo/Dropbox/cursos/longitudinal_data_analysis/longitudinal_data_analysis"
inpfile = "data/lateral_line/JEXBIO-2017-175166-Processed-Data-Master.txt"
fname = file.path(basefolder, inpfile)

danios = fread(fname)
```

Dataset on *giant danios*: how the lateral line system responds to chemical treatments.

The target variables can be:

1.  nearest neighbor distance or NND (unit: body length),
2.  time in school (percentage),
3.  angular bearing (unit: degrees),
4.  angular elevation (unit: degrees),
5.  speed (body length per second).

Explanatory variables include:

-   `Treatment`: gentamycin / sham (control)
-   `Week`: time point in subsequent weeks (from week -1 to week 8)

### EDA

```{r}
danios |>
  group_by(Week, Treatment) |>
  summarise(N = n()) |>
  spread(key = Treatment, value = N) |>
  kable(format = "html", table.attr = "style = \"color: white;\"")
```

```{r, fig.width=12, fig.height=9}
mD <- danios |> gather(key = "target", value = "value", -c(Treatment, Week))
mD$Week <- factor(mD$Week)

df_mean <- mD |> group_by(Week, Treatment, target) |>
  summarise(avg = mean(value))

p <- ggplot(mD, aes(x = Week, y = value)) + geom_boxplot(aes(color = Treatment), alpha = 0.3)
p <- p + geom_point(data = df_mean, mapping = aes(x = Week, y = avg, group=Treatment), color="black")
p <- p + geom_line(data = df_mean, mapping = aes(x = Week, y = avg, group=Treatment, color=Treatment), size = 1.5)
p <- p + facet_wrap(~target, scales = "free_y")
p <- ggpar(p, palette = c("#00AFBB", "#E7B800", "#FC4E07"))
p
```

From reading the article, we expect the treatment to have an effect a little time after the application of gentamycin (vs control/sham) and then to see restoration when the cells of the lateral line system are regenerated.

------------------------------------------------------------------------

**Q: which target variables better show this expected pattern?**

------------------------------------------------------------------------

### Pick target variable

We select `Time in School` (based on the EDA above).

```{r}
dd <- danios |>
  group_by(Week, Treatment) |>
  summarise(avg = mean(`Time in School`), std = sd(`Time in School`))
```

```{r}
temp <- dd |>
  select(-std) |>
  spread(key = Treatment, value = avg) |>
  mutate(diff = abs(Gentamycin-Sham)) |>
  select(-c(Gentamycin,Sham))

dd <- dd |> inner_join(temp, by = "Week")
```

```{r}
datatable(dd, options=list(columnDefs = list(list(visible=FALSE, targets=c(5))),
                           pageLength = 12)) |> 
  # formatStyle('Week', fontWeight = styleInterval(c(2,4), c('normal', 'bold','normal'))) |>
  formatStyle(
    'diff',
    target = 'row',
    fontWeight = styleInterval(c(20), c('normal', 'bold')),
    # color = styleInterval(c(20, 40), c('white', 'blue', 'red')),
    backgroundColor = styleInterval(20, c('white', 'yellow'))
)
```

## Models of analysis

1.  treatment within timepoint
2.  treatment + timepoint
3.  treatment + timepoint + (treatment x timepoint)

### Within timepoint

This is the simplest approach: we split the data by timepoint and make a comparison between treatments.

```{r}
school_time = filter(mD, target == "Time in School") ## !! REMEMBER THAT WE SELECTED ONE TARGET VARIABLE, Time in School !!
school_time$Treatment <- factor(school_time$Treatment, levels = c("Sham", "Gentamycin"))
temp <- filter(school_time, Week == 2)
```

Now we have a much simpler dataset, with `Gentamycin`-treated and control fish records from week 2 only.

$$
\text{Time in School} = \mu + \beta \text{Treatment} + e
$$

```{r}
fit = lm(value ~ Treatment, data = temp)
summary(fit)
```

#### From matrix algebra

$$
\mathbf{y} = \mathbf{Xb} + \mathbf{e}
$$

```{r}
X <- model.matrix(value ~ Treatment, temp)
y = temp$value
```

$$
\mathbf{X'y} = \mathbf{X'Xb}
$$

-   **X**: (n,m) = (50, 2) [50 records, 2 parameters: intercept and slope]
-   **y**: (n,1)
-   **X'y**: (m,1) = (2,1)
-   **X'X**: (m,m) = (2,2)
-   **b**: (m,1) = (2,1)

```{r}
Xy = t(X) %*% y
XX = t(X) %*% X
```

$$
\mathbf{b} = \mathbf{X'X}^{-1} \cdot \mathbf{X'y}
$$

```{r}
b = solve(XX) %*% Xy
print(b)
```

We see that this involves matrix inversion. Since this is a 2x2 matrix, we could do it by hand (for fun! But don't worry: R will take care of matrix inversion for this and -much- larger matrices).

$$
\begin{bmatrix}
a & b \\
c & d
\end{bmatrix} ^ {-1} = \frac{1}{ad-bc} \cdot
\begin{bmatrix}
d & -b \\
-c & a
\end{bmatrix}
$$

```{r}
print(XX)
```

```{r}
multiplicative_factor = 1/(50*25 - 25*25)
M = matrix(c(25, -25, -25, 50), nrow=2, byrow = TRUE)
print(M)
```

```{r}
invMatrix = multiplicative_factor * M
print(invMatrix)
```

```{r}
print(solve(XX))
```

#### And the p-value?

First, we need to estimate the variance of our target variable:

$$
\hat{\sigma}^2 = \frac{1}{(n-2)}\sum(y_i-\hat{y}_i)^2
$$

The $(n-2)$ comes from $(n - (k+1)$, where $k$ is the length of the vector of parameters $\mathbf{b}$ - 1 (to remove the intercept) (we typically look at parameters one by one)

```{r}
n = nrow(temp) ## sample size
y_hat = X %*% b ## predictions/fitted values
residuals = y-y_hat
variance = sum(residuals^2)/(n-2)
print(variance)
```

$$
\text{Var}(\hat{\beta}) = \frac{\hat{\sigma}^2}{\sum(x_i-\overline{x})^2}
$$

```{r}
x_avg = mean(X[,2])
var_beta = variance/sum((X[,2]-x_avg)^2)
```

Now, the standard error of the estimate is the square root of its variance:

```{r}
std_err_beta = sqrt(var_beta)
print(std_err_beta)
```

You can compare this with the results from the `lm()` *R* function:

```{r}
summary(fit)
```

```{r}
tstat = b[2]/std_err_beta
df = n - length(b) ## degrees of freedom
pval = 2*pt(tstat, df = df, lower.tail = TRUE)
print(pval) ## by default a two-tailed test is performed (look at the output from the summary() function!!)
```

#### ANOVA

Alternatively, the same results could be obtained with the Analysis of Variance (ANOVA). More specifically, in this case, one-way ANOVA (only one explanatory variable)

```{r}
ggline(temp, x = "Treatment", y = "value", 
       add = c("mean_se", "jitter"), 
       order = c("Sham", "Gentamycin"),
       ylab = "Time in School", xlab = "Treatment")
```

```{r}
# Compute the analysis of variance
res.aov <- aov(value ~ Treatment, data = temp)
# Summary of the analysis
summary(res.aov)
```

**Anova and linear regression are equivalent!**

#### Apply the within-timepoint analysis to all timepoints

We now take the model used for week 2 and apply it to all weeks (all timepoints): `lm(value ~ Treatment, data = temp)`

```{r}
results = school_time |>
  group_by(Week) |>
  do(tidy(lm(value ~ Treatment, data = .))) |>
  filter(term != "(Intercept)") ## we remove the Intercept estimate and visualize only the effect of treatment
```

```{r}
datatable(results) |> 
  # formatStyle('Week', fontWeight = styleInterval(c(2,4), c('normal', 'bold','normal'))) |>
  formatStyle(
    'p.value',
    target = 'row',
    fontWeight = styleInterval(c(0.05), c('bold', 'normal')),
    # color = styleInterval(c(20, 40), c('white', 'blue', 'red')),
    backgroundColor = styleInterval(0.05, c('yellow', 'white'))
)
```

------------------------------------------------------------------------

### Exercise

Pick another target variable and apply the within-timepoint linear regression analysis:

```{r}
## your code here!
```

------------------------------------------------------------------------

### Across-timepoint analysis

We now use a more complex model of analysis, which uses all the data at once and includes both the effect of timepoint and the effect of treatment:

$$
\text{Time in School} = \mu + \beta_1 \text{Timepoint} + \beta_2 \text{Treatment} + e
$$

Again, we use `Time in School` as target:

```{r}
school_time = filter(mD, target == "Time in School") ## !! REMEMBER THAT WE SELECTED ONE TARGET VARIABLE, Time in School !!
school_time$Treatment <- factor(school_time$Treatment, levels = c("Sham", "Gentamycin"))
```

```{r}
fit = lm(value ~ Treatment + Week, data = school_time)
summary(fit)
```

The output above shows:

-   overall $R^2$ of the model
-   overall p-value of the model
-   estimates of single coefficients (with respect to the reference class)
-   p-values for the single coefficients (under the null hypothesis that they're equal to zero)

------------------------------------------------------------------------

**Q: how do we interpret the model coefficients?**

------------------------------------------------------------------------

We can look also at the global contribution of the two effects included in the model (not the levels of such effects):

```{r}
anova(fit)
```

Specific contrasts between levels of the categorical explanatory variables can be obtained with a number of ways. E.g., buy using the `TukeyHSD()` function (e.g. [here](https://en.wikipedia.org/wiki/Tukey%27s_range_test)) which requires an `aov` object:

```{r}
res.aov = aov(value ~ Week + Treatment, data = school_time)
TukeyHSD(res.aov)
```

Alternatively, the `contrast` *R* package provides the `contrast()` function:

```{r}
contrast(fit, 
         list(Week = "2", Treatment = "Sham"),
         list(Week = "2", Treatment = "Gentamycin"))
```

```{r}
contrast(fit, 
         list(Week = "-1", Treatment = "Sham"),
         list(Week = "8", Treatment = "Sham"))
```

In this notebook, we don't delve much into details on contrasts, though.

### Interaction model

Finally, we can include in the model also the interaction between timepoint and treatment:

$$
\text{Time in School} = \mu + \beta_1 \text{Timepoint} + \beta_2 \text{Treatment} + \beta_3 (\text{Timepoint} * \text{Treatment}) + e
$$

```{r}
fit = lm(value ~ Week + Treatment + Week:Treatment, data=school_time)
summary(fit)
```

- `Week0`: average `Time in School` at Week 0 vs Week -1, irrespective of treatment and interaction
- `Gentamycin`: average `Time in School` of treated fish, irrespective of time and interaction
- `Week0:Gentamycin`: average difference in the effect of `Gentamycin` on `Time in School` between `Week0` and `Week -1`

Fish at week 1, treated with gentamycin

$$
98.6 + (-16.12) + (0.5875) + (-28.4675) = 54.6
$$

From the across-timepoints model:

$$
104.862 + (-15.46) + (-27.902) = 61.5
$$

```{r}
98.6 + (-16.12) + (0.5875) + (-28.4675)
```

Raw average $y$ at week 1 for treated fish = **54.6** !

```{r}

anova(fit)
```

------------------------------------------------------------------------

**Q: is the additivity assumption valid?** **Q: has model fit improved with the interaction term?** **Q: how do we interpret the interaction terms?**

------------------------------------------------------------------------
