---
title: 'Timepoints and treatments: exercise'
author: "Filippo Biscarini"
date: "2023-06-19"
output: html_document
---

```{r setup, include=FALSE}
library("DT")
library("broom")
library("knitr")
library("ggpubr")
library("contrast")
library("tidyverse")
library("data.table")

knitr::opts_chunk$set(echo = TRUE)
```


## Read the data

Data from [The early bird gets the worm: foraging strategies of wild songbirds lead to the early discovery of food sources](https://royalsocietypublishing.org/doi/10.1098/rsbl.2013.0578)
(data repo [here](https://zenodo.org/record/4936819))

It's data on time needed to discover food sources in songbirds.

```{r read-the-bird-data}
basefolder = "/home/filippo/Dropbox/cursos/longitudinal_data_analysis/longitudinal_data_analysis"
inpfile = "data/bird_food/discovery_individuals.csv"
fname = file.path(basefolder, inpfile)

birds = fread(fname)
```

Dataset on wild songbirds and their foraging habits (from here: "The early bird gets the worm", https://zenodo.org/record/4936819).

The target variable is *time to discovery* (of food: continuous, in seconds).
Explanatory variables include:

- `TREATMENT`: morning / afternoon
- `HALFHOUR`: time during the day
- `SITE`: geographic place
- `WEEK`: replicates over two weeks


We can consider *week* to be a replicate of the experiment (two different weeks, like two random replicates),
and we can combine the part of the day (morning/afternoon) and half hour to have a unique expression of the time of the day.

This will lead to the following model.

$$
\text{TIME.TO.DISC} = \mu + time + \beta_1 \text{time} + \beta_2 \text{SITE} + e
$$

### Task 1: preprocessing - modify the dataset

```{r}
## your code here!
```

### Task 2: EDA

```{r}
## your code here!
```

### Task 3: choose and fit your model

```{r}
## your code here!
```

### Task 4: visualize and interpret results

```{r}
## your code here!
```



