---
title: "tidymodels"
author: "Filippo Biscarini"
date: "2023-06-21"
output: html_document
---

```{r setup, include=FALSE}
library("DT")
library("broom")
library("knitr")
library("ggpubr")
library("contrast")
library("tidyverse")
library("tidymodels")
library("data.table")

knitr::opts_chunk$set(echo = TRUE)
```

A gentle introduction to `tidymodels`

## Read the data

Data from [The effects of lateral line ablation and regeneration in schooling giant danios](https://journals.biologists.com/jeb/article/221/8/jeb175166/300/The-effects-of-lateral-line-ablation-and) (data repo [here](https://zenodo.org/record/4999506))

It's data on fish (*Devario aequipinnatus*) lateral line system, with the effect of chemical treatments (gentamycin or no treatment -- sham) at different timepoints.

```{r read-the-bird-data}
basefolder = "/home/filippo/Dropbox/cursos/longitudinal_data_analysis/longitudinal_data_analysis"
inpfile = "data/lateral_line/JEXBIO-2017-175166-Processed-Data-Master.txt"
fname = file.path(basefolder, inpfile)

danios = fread(fname)
```

Dataset on *giant danios*: how the lateral line system responds to chemical treatments.

Variables include:

* nearest neighbor distance or NND (unit: body length),
* time in school (percentage),
* angular bearing (unit: degrees),
* angular elevation (unit: degrees),
* speed (body length per second).
-   `Treatment`: gentamycin / sham (control)
-   `Week`: time point in subsequent weeks (from week -1 to week 8)

A little preprocessing:

```{r}
danios <- danios |>
  mutate(Treatment = factor(Treatment, levels = c("Sham", "Gentamycin")))
```


Let's say that we want to study the correlation between two such variables, e.g. `NND` and `Time in School`.
We can start from the entire dataset:

```{r}
ggplot(danios, aes(`Time in School`, NND)) + geom_jitter()
```


```{r}
cor(danios$`Time in School`, danios$NND)
```

We see that overall there is a moderately negative relationship between the time spent in school and the distance between individuals (linear correlation coefficient = -0.43). 
We also notice that there's a concentration of data points towards the maximum recorded value for time in school (100% time spent in school)

We can look at the relationship between NND and Time in School stratifying by week (time) and treatment: `dplyr` (from `tidyverse`) gives us a nice way to do it.

```{r}
ggplot(danios, aes(`Time in School`, NND, color = Treatment)) +
  geom_point() + facet_wrap(~Week, scales = "free_y")
```

```{r}
danios |>
  group_by(Week, Treatment) |>
  summarise(correlation = cor(`Time in School`, NND))
```

What if we want to make a hypothesis test on the correlation coefficients?
In this case the output is not a single number (as was the case with `cor`), but a complex object with vectors, strings, numbers, lists etc.

```{r}
ct = cor.test(danios$`Time in School`, danios$NND)
ct
```

We can tidy it! (`broom` inside)

```{r}
broom::tidy(ct)
```

**Why `tidymodels`?**

- perform multiple tests
- fit multiple models 
- on different subsets of the data

The `nest-map-unnest` workflow is what we need. 

Example: correlation tests for each Week and Treatment. 

#### 1. Nesting

We start by nesting our data based on the group -or groups- of interest:
we need to pick the variables that are "nested" within group(s) (e.g. NND and Time in School nested within Week and Treatment).

```{r}
nested <- 
  danios %>% 
  select(NND, `Time in School`, Treatment, Week) |>
  nest(data = c(NND, `Time in School`))

nested
```

#### 2. Perform the analysis

Then we perform the correlation test for each nested subset using `purrr::map()`:

```{r}
nested <- nested %>% 
  mutate(test = map(data, ~ cor.test(.x$`Time in School`, .x$NND)))

nested
```

#### 3. Tidy-up results

```{r}
nested <- nested |>
  mutate(tidied = map(test, tidy))

nested
```

#### 4. Unnest and visualize

```{r}
nested |>
  unnest(cols = tidied) |>
  select(-data, -test)
```

We could of course put all the steps together in a single workflow: **dare to try?**

```{r}
## your code here
```


### Tidy regression

The output of linear regression (function `lm()`) is well known to be particularly "untidy".
Let's take the example of Time in school as a function of treatment (gentamycin, control):

```{r}
lm_fit <- lm(`Time in School` ~ Treatment, data = danios)
summary(lm_fit)
```

We can tidy it:

```{r}
broom::tidy(lm_fit)
```

And use `tidymodels` for the within-week regression analysis:

```{r}
danios %>%
  select(`Time in School`, Treatment, Week) |>
  nest(data = c(-Week)) %>% 
  mutate(
    fit = map(data, ~ lm(`Time in School` ~ Treatment, data = .x)),
    tidied = map(fit, tidy)
  ) %>% 
  unnest(tidied) %>% 
  select(-data, -fit)
```

We can use multiple `broom` functions at a time and save results:

- `tidy`: tidies up results, e.g. estimates of coefficients and p-values
- `glance`: returns overall goodness of fit measures of the model
- `augment`: returns fitted values (predictions) and residuals, for model evaluation and model diagnostics

```{r}
regressions <- 
  danios %>%
  select(`Time in School`, Treatment, Week) |>
  nest(data = c(-Week)) %>% 
  mutate(
    fit = map(data, ~ lm(`Time in School` ~ Treatment, data = .x)),
    tidied = map(fit, tidy),
    glanced = map(fit, glance),
    augmented = map(fit, augment)
) 
```

```{r}
regressions %>% 
  select(tidied) %>% 
  unnest(tidied)
```

```{r}
regressions %>% 
  select(glanced) %>% 
  unnest(glanced)
```

```{r}
regressions %>% 
  select(augmented) %>% 
  unnest(augmented)
```
