---
title: "Difference-in-differences"
author: "Filippo Biscarini"
date: "2023-06-20"
output: html_document
---

```{r setup, include=FALSE}
library("plm")
library("tidyverse")
library("data.table")

knitr::opts_chunk$set(echo = TRUE)
```

 ## A little background

Differences in the outcome variable (Y), i.e. a before-after comparison (e.g. treatment vs control):

- observations from regions that underwent intervention vs. observations from regions where the intervention did not take place
  - treatment = intervention
  - time = before/after
- observations from a region before and after the intervention
  - treatment = intervention
  - synthetic control needed

**The main assumption is that without the change in the environment (intervention/policy etc.) the outcome variable would have remained constant!**

The **diff-in-diff** approach includes a before-after comparison for a **treatment** and **control** group. 
This is a combination of:

- a *cross-sectional comparison* (treated vs non-treated control group)
- a **before-after** (*longitudinal*) **comparison** (treatment group with itself, before and after the treatment)

The before-after difference in the treatment group gets a correction, by accounting for the before-after difference in the control group, eliminating the trend problem. 

To obtain an unbiased estimate of the treatment effect one needs to make a parallel trend assumption. 
That means without the change in the environment, 
the change in the outcome variable would have been the same for the two groups (**counterfactual** outcome).

The validity of the diff-in-diff approach is closely related to the similarity of the treatment and control groups. 
Hence, some plausibility checks should be conducted:

- compute placebo-diff-in-diff for periods without a change in the environment
- for (longer) time series: check and demonstrate the parallel time trends
- use an alternative control group (if available, or synthetic): the estimate should be the same
- replace **y** by an alternative outcome which is known to be independent of the treatment (the diff-in-diff estimator should be 0)

## Diff-in-diff by hand

First, we walk through the calculations withoout using an explicit model.

### Injury dataset

- **1980**: new policy (Kentucky) on raised weekly earnings that were covered by worker's compensation (more after-injury benefits for high-earners). 
- **Research question**: has this new policy caused higher earning workers to spend more time injured? (generous benefits may cause higher earning workers to be more reckless on the job, or to claim that off-the-job injuries were incurred while at work, or to prefer injury benefits rather than keep on working - for mild injuries at least)

```{r injury}
basefolder = "/home/filippo/Dropbox/cursos/longitudinal_data_analysis/data"
fname = file.path(basefolder, "injury/injury.csv")
injury <- read_csv(fname)
```

### Preprocessing

Rename columns:

```{r}
injury <- injury %>% 
  filter(ky == 1) |>
  rename(duration = durat, log_duration = ldurat,
         after_1980 = afchnge,
         earnings = highearn) ## kentucky is where the policy was implemented, other states are controls
```


- `duration`: duration of injury benefits, measured in weeks
- `log_duration`: `log(duration)` [natural logarithm]
- `after_1980`: observation happened before (0) or after (1) the policy change in 1980. This is our time (or before/after variable)
<!-- - `policy`: states that implemented (Kentucky, `1`) or not (other states, `0`) the policy on unemployment benefits -->
- `highearn`: observation is a low (0) or high (1) earner. This is our group (or treatment/control) variable: there was no change for low earners (same benefits), while higher earners now have benefits that they did not have earlier 

```{r}
injury <- injury |>
  mutate(earnings = ifelse(earnings == 0, "low-income","high-income"),
         after_1980 = ifelse(after_1980 == 0, "before", "after"))
```


```{r}
injury <- injury |>
  mutate(earnings = factor(earnings, levels = c("low-income","high-income")),
         after_1980 = factor(after_1980, levels = c("before", "after")))
```

### EDA

```{r}
ggplot(data = injury, aes(x = duration)) +
  # binwidth = 8 makes each column represent 2 months (8 weeks) 
  # boundary = 0 make it so the 0-8 bar starts at 0 and isn't -4 to 4
  geom_histogram(binwidth = 8, color = "white", boundary = 0) +
  facet_wrap(vars(earnings)) ## groups
```

The distribution is really skewed, with most persons in both groups getting the lowest range of benefits weeks.
Using the logarithm of duration would change this, making the distribution of the $y$ more "gaussian", hence more amenable
to be analysed with linear regression models.

```{r}
ggplot(data = injury, mapping = aes(x = log_duration)) +
  geom_histogram(binwidth = 0.5, color = "white", boundary = 0) + 
  # Uncomment this line if you want to exponentiate the logged values on the
  # x-axis. Instead of showing 1, 2, 3, etc., it'll show e^1, e^2, e^3, etc. and
  # make the labels more human readable
  # scale_x_continuous(labels = trans_format("exp", format = round)) +
  facet_wrap(vars(earnings))
```


Let's plot average log-durations in the two groups, before and after the implementation of the policy:
we see that higher-income workers had already a higher n. of injury-benefits weeks before the new policy 
(maybe workers that do riskier jobs are paid better).
With the new policy, this seems emphasized.

```{r}
ggplot(injury, aes(x = earnings, y = log_duration)) +
  geom_point(size = 0.5, alpha = 0.2) +
  stat_summary(geom = "point", fun = "mean", size = 5, color = "red") +
  facet_wrap(vars(after_1980)) + xlab("policy")
```

```{r}
plot_data <- injury %>% 
  group_by(earnings, after_1980) %>% 
  summarize(mean_duration = mean(log_duration),
            se_duration = sd(log_duration) / sqrt(n()),
            upper = mean_duration + (1.96 * se_duration),
            lower = mean_duration + (-1.96 * se_duration)) 
```

```{r}
ggplot(plot_data, aes(x = earnings, y = mean_duration)) +
  geom_pointrange(aes(ymin = lower, ymax = upper), 
                  color = "darkgreen", size = 1) +
  facet_wrap(vars(after_1980)) + xlab("policy") + ylab("mean duration (log scale)")
```

We can now see the change, in terms of number of weeks of injury benefits (log scale) before and after the new policy,
in the two groups:

```{r}
ggplot(plot_data, aes(x = after_1980, y = mean_duration, color = earnings)) +
  geom_pointrange(aes(ymin = lower, ymax = upper), size = 1) + 
  # The group = highearn here makes it so the lines go across categories
  geom_line(aes(group = earnings)) + ylab("mean duration (log scale)")
```

## Diff-in-diff by hand

After having explored the data, we can now actually calculate the estimate of the **difference in differences** for the two groups:
difference between after-before differences in high vs low income workers.

```{r}
diffs <- injury %>% 
  group_by(after_1980, earnings) %>% 
  summarize(mean_duration_log = mean(log_duration),
            # Calculate average with regular duration too, just for fun
            mean_duration_weeks = mean(duration))
diffs
```

```{r}
dd <- diffs |>
  select(-mean_duration_weeks) |>
  spread(key = after_1980, value = mean_duration_log) |>
  mutate(diff = after-before) |>
  mutate(across(where(is.numeric), round, 3))

last_row = cbind.data.frame(data.frame("earnings"="diff"),diff(as.matrix(dd[,-1])))
dd <- bind_rows(dd,last_row)
print(dd)
```

The **diff-in-diff estimate** is 0.19, which means that the program causes an increase in unemployment duration of 0.19 logged weeks.
For *log-linear models* ($log(y) = \mu + \beta x + e$), this translates to $e^{0.19}=1.2$ weeks.

This is shown graphically in the plot below:

```{r}
ggplot(diffs, aes(x = after_1980, 
                  y = mean_duration_log, 
                  color = earnings)) + 
  geom_point() +
  geom_line(aes(group = earnings)) +
  # If you use these lines you'll get some extra annotation lines and
  # labels. The annotate() function lets you put stuff on a ggplot that's not
  # part of a dataset. Normally with geom_line, geom_point, etc., you have to
  # plot data that is in columns. With annotate() you can specify your own x and
  # y values.
  annotate(geom = "segment", x = "before", xend = "after",
           y = pull(dd[2,2]), yend = pull(dd[2,3]) - pull(dd[3,4]),
           linetype = "dashed", color = "grey50") +
  annotate(geom = "segment", x = "after", xend = "after",
           y = pull(dd[2,3]), yend = pull(dd[2,3]) - pull(dd[3,4]),
           linetype = "dotted", color = "blue") +
  annotate(geom = "label", x = "after", y = pull(dd[2,3]) - (pull(dd[3,4]) / 2), 
           label = "Program effect", size = 3)
```

