---
title: "Survival analysis"
author: "Filippo Biscarini"
date: "21/06/2023"
output: html_document
---

```{r setup, include=FALSE}
library("knitr")
library("tidyverse")

knitr::opts_chunk$set(echo = TRUE)
```

## A primer on survival analysis

## Cox models

The Cox proportional-hazards model (Cox, 1972) is essentially a regression model commonly used in medical research for investigating the association between the survival time of patients and one or more predictor variables.

### Time-to-event data

**time-to-event data (survival analysis)**:

Examples of types of events:

-   Relapse
-   Progression
-   Death

Characteristics of time events:

-   subjects enter at different times and have different duration of follow-up
-   entire survival experience, not just the percentages who remain alive at the end of the study
-   the survival distributions may differ even though the five-year survival rates are similar

## Survival analysis

```{r}
library("survival")
library("survminer")

# data("lung")
head(lung) |> kable(format = "html", table.attr = "style = \"color: white;\"")
```

-   `time`: days
-   `status`: `1` = censored; `2` = dead
-   `sex`: 1 = male; 2 = female

```{r}
lung |> 
  group_by(sex, status) |> 
  summarise(N = n()) |>
  spread(key = status, value = N)
```

-   a survival object created using the function `Surv()`: survival time (time from enrollment to end of study), status at the end of study (dead/censored)
-   the data set containing the variables: covariates (e.g. sex)

```{r}
fit <- survfit(Surv(time, status) ~ sex, data = lung)
print(fit)
```

```{r, label='kaplar-meier-curve', fig.height=10, fig.width=10}
# Change color, linetype by strata, risk.table color by strata
ggsurvplot(fit,
          pval = TRUE, conf.int = TRUE,
          pval.method = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ncensor.plot = TRUE,
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"))
```

```{r}
ggsurvplot(fit,
          conf.int = TRUE,
          risk.table.col = "strata", # Change risk table color by groups
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"),
          fun = "event")
```

#### Cummulative hazard

It is defined as $H(t)=−log(survivalfunction)=−log(S(t))$

The cumulative hazard ($H(t)$) corresponds to the number of events that would be expected for each individual by time `t` if the event were a repeatable process

```{r}
ggsurvplot(fit,
          conf.int = TRUE,
          risk.table.col = "strata", # Change risk table color by groups
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"),
          fun = "cumhaz")
```

#### Kaplan-Meier life table: summary of survival curves

-   time: the time point at which the curve has a step
-   n.risk: the number of subjects at risk at time *t*
-   n.event: the number of events that occur at time *t*
-   n.censor: number of censored events at time *t*
-   surv: estimate of survival

```{r}
res.sum <- surv_summary(fit)
head(res.sum, 10) |> kable(format = "html", table.attr = "style = \"color: white;\"")
```

#### Log-Rank test comparing survival curves: survdiff()

Tests if there is a difference between two or more survival curves: `rho` = 0 is log-rank test

```{r}
surv_diff <- survdiff(Surv(time, status) ~ sex, data = lung, rho = 0)
surv_diff
```

## From Kaplan-Meier curves to Cox models

Kaplan-Meier curves and log-rank tests are useful for univariate analysis, describing survival in terms of one factor under investigation, and typically work only with categorical predictors (e.g. sex, treatment A vs treatment B etc.)

This is where **Cox proportional hazards regression analysis** comes in handy: it works for both quantitative predictor variables and for categorical variables. Furthermore, the Cox regression model extends survival analysis methods to assess simultaneously the effect of several risk factors on survival time.

Cox models examine how specified factors (**covariates**) influence the rate (**hazard rate**) of a particular event happening (e.g., infection, death) at a particular point in time

The Cox model is expressed by the **hazard function** denoted by $h(t)$, interpretable as the risk of dying at time `t`:

$$
h(t)=h_0(t) \cdot exp(\beta_1x_1 + \beta_2x_2 + \ldots + beta_kx_k)
$$

```{r}
res.cox <- coxph(Surv(time, status) ~ sex, data = lung)
res.cox
```

#### Multiple Cox regression model

```{r}
res.cox <- coxph(Surv(time, status) ~ age + sex + ph.ecog, data =  lung)
summary(res.cox)
```

```{r}
library("lmtest")
lrtest(res.cox,"ph.ecog")
```
