---
title: "Time series and autocorrelations"
author: "Filippo Biscarini"
date: "2023-06-22"
output: html_document
---

```{r setup, include=FALSE}

library("ggplot2")
library("tseries")
library("forecast")
library("tidyverse")

knitr::opts_chunk$set(echo = TRUE)
```

Autocorrelation is used to measure the degree of similarity between a time series and a lagged version of itself over the given range of time intervals. 
We can also call autocorrelation as *serial correlation* or *lagged correlation*. 
It is mainly used to measure the relationship between the actual values and the previous values.

Data on Mauna Loa monthly atmospheric $\text{CO}_2$ concentration

```{r co2}
data(co2)
co2
```

```{r}
plot(co2)
```


```{r}
co2_df = data.frame("CO2" = co2)
co2_df$year = rep(seq(1959,1997), each=12)
co2_df$month = rep(month.abb, times=(1997-1959+1))

co2_df <- co2_df |>
  mutate(month = factor(month, levels = c(month.abb)),
         year = factor(year),
         CO2 = as.numeric(CO2))
```

```{r}
ggplot(co2_df, aes(x = month, y = CO2, group=year)) + geom_line(aes(color=year))
```

$$
r_k = \frac{\sum_{t=k+1}^T (y_t-\bar{y})(y_{t-k}-\bar{y})}{\sum_{t=1}^T(y_t-\bar{y})^2}
$$

```{r}
v = c(12,34,56,30,23,15,28,36,45,31,21,10,29,40,58,64,78,90)
plot(v)
```


With lag = 0, the autocorrelation is clearly $1$: $\frac{\sum (y_t-\bar{y}) \cdot (y_t-\bar{y})}{\sum (y_t-\bar{y})^2}$.

With lag = $1$, we calculate the correlation between the two vectors ($n$ is the length of the vector):

- v[2,n]
- v[1,n-1]

```{r}
n = length(v)
v_avg = mean(v)
v1 = v[2:n]
v2 = v[1:(n-1)]

## autocorrelation with lag = 1
r1 = ((v1 - v_avg) %*% (v2 - v_avg))/((v - v_avg) %*% (v - v_avg))
print(r1)
```

**Q: can you calculate the autocorrelation with lag = 2?** (let's do this together!)

```{r}
## your code here!!
```



The `acf()` function from the `stats` package:

- input vector (time series or spatial series etc.)
- lag

```{r}
acf(v, plot = FALSE)
```

```{r}
data(ausbeer)
beer2 <- window(ausbeer, start=1992)
```

```{r}
acf(beer2, plot = FALSE)
```

```{r}
ggAcf(beer2)
```


```{r}
acf(co2, plot = FALSE)
```


