---
title: "Time series and autocorrelations"
author: "Filippo Biscarini"
date: "2023-06-22"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}

library("fpp")
library("fpp2")
library("GGally")
library("ggplot2")
library("tseries")
library("corrplot")
library("forecast")
library("tidyverse")
library("data.table")

knitr::opts_chunk$set(echo = TRUE)
```

## Time series

A time series is a list of numbers ordered by time.
Time series-like objects can be also other types of ordered data: e.g. spatial series (geographical proximity), genome sequences (e.g. adjacent loci/nucelotides).
More generally, these data are known as **sequence data** (text, sounds are other types of sequence data).

In R, time series data can be stored as a `ts` object.

Data on Mauna Loa monthly atmospheric $\text{CO}_2$ concentration

```{r co2}
data(co2)
class(co2)
```

```{r}
co2
```

```{r}
autoplot(co2) + ggtitle("Monthly atmospheric CO2 concentration in Manua Loa")
```

Convert time series object to dataframe:

```{r}
co2_df = data.frame("CO2" = co2)
co2_df$year = rep(seq(1959,1997), each=12)
co2_df$month = rep(month.abb, times=(1997-1959+1))

co2_df <- co2_df |>
  mutate(month = factor(month, levels = c(month.abb)),
         year = factor(year),
         CO2 = as.numeric(CO2))
```

```{r}
ggplot(co2_df, aes(x = month, y = CO2, group=year)) + geom_line(aes(color=year))
```

```{r}
ggseasonplot(co2, year.labels=TRUE, year.labels.left=TRUE) +
  ylab("CO2") +
  ggtitle("Seasonal plot: atmospheric CO2 concentration")
```

```{r}
ggseasonplot(co2, polar=TRUE) +
  ylab("CO2") +
  ggtitle("Polar seasonal plot: atmospheric CO2 cincentration")
```

```{r}
ggsubseriesplot(co2) +
  ylab("CO2") +
  ggtitle("Subseries plot: monthly CO2 concentration in the atmosphere over the 1959-1997 period")
```

The horizontal lines indicate the means for each month.
This plot highlights the underlying seasonal pattern to be seen clearly, and also shows the changes in seasonality over time.

Now we use data from two paired time series: electricity demand (Gigawatts) and temperature (Celsius degrees) from Victoria (Australia) in 2014:

```{r}
data(elecdemand) ## from fpp2
head(elecdemand)
```

```{r}
autoplot(elecdemand[,c("Demand","Temperature")], facets=TRUE) +
  xlab("Year: 2014") + ylab("") +
  ggtitle("Half-hourly electricity demand: Victoria, Australia")
```

We can use a scatterplot to look at the relationship between these two time series:

```{r}
df <- as.data.frame(elecdemand)
ggplot(df, aes(Temperature, Demand)) + geom_point()
```

```{r}
co2x <- window(co2, start=1980)
gglagplot(co2x, lags = 6)
```

#### European stillbirth data

European Stillbirth Rate Time Series Dataset (data repo at: <https://zenodo.org/record/6505519>)

```{r read-stillbirth-data}
basefolder = "/home/filippo/Dropbox/cursos/longitudinal_data_analysis/longitudinal_data_analysis"
inpfile = "data/stillbirth/sbr_all.xlsx"
fname = file.path(basefolder, inpfile)

stillbirth = readxl::read_xlsx(fname, sheet = 2)
```

```{r}
temp = ts(stillbirth[,-1], start = 1775, end = 2021, names = names(stillbirth)[-1])
```

```{r}
autoplot(temp)
```

```{r, warning=FALSE, message=FALSE}
GGally::ggpairs(as.data.frame(temp[,-3]))
```

## Autocorrelations

Autocorrelation is used to measure the degree of similarity between a time series and a lagged version of itself over the given range of time intervals.
We can also call autocorrelation as *serial correlation* or *lagged correlation*.
It is mainly used to measure the relationship between the current values and the previous values.

$$
r_k = \frac{\sum_{t=k+1}^T (y_t-\bar{y})(y_{t-k}-\bar{y})}{\sum_{t=1}^T(y_t-\bar{y})^2}
$$

```{r}
v = c(12,34,56,30,23,15,28,36,45,31,21,10,29,40,58,64,78,90)
plot(v)
```

With lag = 0, the autocorrelation is clearly $1$: $\frac{\sum (y_t-\bar{y}) \cdot (y_t-\bar{y})}{\sum (y_t-\bar{y})^2}$.

With lag = $1$, we calculate the correlation between the two vectors ($n$ is the length of the vector):

-   v[2,n]
-   v[1,n-1]

```{r}
n = length(v)
v_avg = mean(v)
v1 = v[2:n]
v2 = v[1:(n-1)]

## autocorrelation with lag = 1
r1 = ((v1 - v_avg) %*% (v2 - v_avg))/((v - v_avg) %*% (v - v_avg))
print(r1)
```

**Q: can you calculate the autocorrelation with lag = 2?** (let's do this together!)

```{r}
## your code here!!
```

The `acf()` function from the `stats` package:

-   input vector (time series or spatial series etc.)
-   lag

```{r}
acf(v, plot = FALSE)
```

```{r}
forecast::ggAcf(ts(v))
```

```{r}
acf(co2, lag = 2, plot = FALSE)
```

```{r}
acf(co2, lag.max = 6, plot = FALSE)
```

```{r}
ggAcf(co2)
```

```{r}
time(elecdemand)[17000]
elecdemand2 <- window(elecdemand, start = 2014.95)
ggAcf(elecdemand2[,"Demand"])
```

```{r}
ggAcf(elecdemand2[,"Temperature"])
```

```{r}
ggAcf(temp[,7])
```

### White noise time series

Time series that show no autocorrelation are called white noise.
Figure 2.17 gives an example of a white noise series.

```{r}
y <- ts(rnorm(50))
autoplot(y) + ggtitle("White noise")
```

```{r}
ggAcf(y)
```

For **white noise series**, we expect each autocorrelation to be close to zero.
Of course, they will not be exactly equal to zero as there is random variation.
For a white noise series, we expect $95\%$ of the spikes in the autocorrelation plot to lie within:

$$ 
\pm \frac{2}{\sqrt{T}} 
$$

where $T$ is the length of the time series.

It is common to plot these bounds on the graph (the blue dashed lines above).
If one or more large spikes are outside these bounds, or if substantially more than 5% of spikes are outside these bounds, then the series is probably not white noise.

In this white noise series, $T = 50$ and so the bounds are at $\pm \frac{2}{\sqrt{50}}= \pm 0.28$.
All of the autocorrelation coefficients lie within these limits, confirming that the data are white noise.
