---
title: "Analysis_outliers and influencers"
output: html_document
date: "2023-06-29"
author: "Andreia J. Amaral"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(car)
```

## Identifying outliers and influencer values

In this R notebook we discuss metrics and R functions for the detection of outlier and influencer data points

There are three types of "ways" in which data points my stand out from the remaining:

- Leverage points: a data point that is very different in its predictor behavior than the rest if the data. In terms of regression it means that it’s x-value is much higher or lower than the predictor mean

-Discrepancy points – When a point has unsual Y-value in comparison with the X-value. This is an outlier in the regression framework

-Influencer points – The influence of a single data is determined by the relation Leverage x Discrepancy. An influencer point has high discrepancy and high leverage

## Read data and Measure Leverage

The first assessment that we will make is of "LEVERAGE"
For that we start by reading the following data set diagnosts_examples.txt

The diagnostic measure of Leverage is the HAT VALUE, which is the distance to the mean of the predictor (mean(X)) for each individual predictor point (Xi). The standardized version of this distance is called the hat-value.

```{r}

toy_data<-read.table("/Users/andreiafonseca/Documents/Curso_Physalia_Longitudinal_data/diagnostics_examples.txt", header=T, sep="\t")

toy_data


# distance to mean of point 8
numerator = (toy_data$Nclinics[8] - mean(toy_data$Nclinics))**2

# distance to mean of all the other points
denominator = sum((toy_data$Nclinics - mean(toy_data$Nclinics))**2)
n=8
# putting it together
h_8 = 1/n+numerator/denominator

h_8 # h_8=0.969 which very high very close to 1 the maximum value

# fitting linear model
mod_surviving = lm(Surviving~Nclinics, data=toy_data)

# getting hatvalues and printing them
hs = hatvalues(mod_surviving)
as.numeric(hs)
```
## Assessing discrepancy 

Points that do not have a good fit to the linear regression line are considered outliers or are points with high discrepancy
Diagnostic measure: studentized residuals

Studentized residuals are calculated by fitting a model without the case for which the residual is calculated, and then scaling the resulting residual (Ei) by an estimate of the standard deviation of the residuals (SE(−i)) and the point's hat value.

E*i= Ei/SE(-i)* sqrt(1-hi)


```{r}
##assessing discrepancy

# Extract the residuals from the linear regression model
residuals <- resid(mod_surviving)

# Plot Clinics vs Residuals
plot(toy_data$Nclinics, residuals,
     xlab = "Clinics", ylab = "Residuals",
     main = "Clinics vs Residuals", pch = 16)

# Calculate studentized residuals using rstudent function
student_resid <- rstudent(mod_surviving)

# Plot Clinics vs Studentized Residuals
plot(toy_data$Nclinics, student_resid,
     xlab = "Clinics", ylab = "Studentized Residuals",
     main = "Clinics vs Studentized Residuals", pch = 16)
```

##Assessing influence

Points with high leverage and high discrepancy have high influence

Diagnostic measures are:
    1. Cook’s d
    2. DFFITS: scales the studentized residual by the corresponding hat value
Both require the hat value and the studentized resildual

```{r}
##accessing influence

#estimate DFFIT for all data points
as.numeric(dffits(mod_surviving))

#estimate COOKS DISTANCE for all data points
as.numeric(cooks.distance(mod_surviving))

# plot these parameters and visualize
influencePlot(mod_surviving)

# Calculate hat values and studentized residuals
hat_values <- hatvalues(mod_surviving)
student_resid <- rstudent(mod_surviving)

# Create a data frame with hat values and studentized residuals
df <- data.frame(Hat_Values = hat_values, Studentized_Residuals = student_resid)

# Create the scatter plot using ggplot and plt hat values and studentized residuals
ggplot(df, aes(x = Hat_Values, y = Studentized_Residuals)) +
  geom_point() +
  xlab("Hat Values") +
  ylab("Studentized Residuals") +
  ggtitle("Hat Values vs Studentized Residuals")



```

## Exercise

what about checking these parameters to your data?

```{r}
###put your code here
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
